The Evolution of Phenotypically Invariant Genetic Networks

### 2017-09-27 kalman_walk_paper.tex
edited kalman walk and figure 6

\newcommand{\Sys}{\mathcal{S}}
\newcommand{\allS}{\mathcal{K}}
%\newcommand{\allS}{\mathcal{A}}

@@

\plr{Additional ideas to consider adding:}
\begin{itemize}
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    \item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
\end{itemize}

\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}

\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}
Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
  \plr{Saying "stalled" reflects negatively on other fields - rephrase?} \js{I don't think it reflects negatively. Even if it did, as long as it's a true statement it should be fine.}.
There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
  \plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.} \js{deleted HW. I don't think we need to write an anodyne -- the current sentence is true enough.}
However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
This outlook necessitates the advancement of conceptual frameworks of such precision, 
only mathematics will suffice, as models allow the development of concrete numerical predictions. 
%Previously it has been suggested that any idealized study of evolution is incomplete 
%without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.
\plr{probably some more recent Wagner papers in this line as well?}
%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory. 
%
%Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
%
%
%It is commonly taught that an organism's genome contains the heritable material 
%that natural selection filters and that an organism's phenotype directly determines its
%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
%undrstood molecular machinery -- and it is a major goal common to many different disciplines
%within the life sciences to elucidate its form, function, and evolution. These aims are
%delicately intertwined and a comphrehensive understanding of a system's evolution requires
%an understanding of its function and vice versa.

@@

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
%\plr{not clear what you mean by "actually"}\js{I think we discussed it's clear enough for an intro?}
Models with different parameter schemes, or even different mechanics can be equally accurate, 
but still not \emph{actually} agree with what is being modelled. 
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
In chemical engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.  
\plr{here you are jumping to evolution -- haven't said yet why evolution "can explore" these -- needs to say explicitly what we mean by neutral here}
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; 
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

In this paper we outline a theoretical framework to study the evolution of biological systems. 
  \plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}\js{fixed}
Presently, we focus solely on a neutral scenario, 
that is where phenotype is conserved over evolutionary time. 
however, this framework could be applied to a wider set of evolutionary scenarios. 
\plr{here's the definition of neutral, but it's not quite right?} \js{how's that?}

@@

\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".} \js{added the word ``linear.''}
-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all linear biological systems can, in principal, can undergo systems drift. 
In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}

@@

Such a biological system $\mathcal{S}$ is given by,
  \begin{align}
    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}

@@

%\plr{This paragraph feels like the introduction.}
%The literature is filled with detailed observations of molecular systems and their diversity. 
%\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
%There are examples of significant diversity in the networks underlying processes such as 
%circadian rhythm \citep{sancar2008intelligent}, 
%cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
%pattern formation \plr{cite?}, 
%and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
Despite a symmetry in functionality or phenotype systems can often differ, sometimes substantially, at the molecular level. 
How many different mechanisms have the same function?

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal gene network 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

%Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
%That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
%\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
%    i.e. behaviour across many possible inputs.}
%Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
%  \begin{align}
%    VB &= B \implies \bar{B} = B \\
%    CV &= C \implies \bar{C} = C
%  \end{align}
%\plr{this sentence lacks an object (or some sentence part?)}
%In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VAV^{-1}, VB, CV )$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
  \begin{align}
    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
  \js{what do you like better for $\allS$? $\mathcal{A}$, $\mathcal{A}_{\infty}$, or $\mathcal{K}_{h(t)}$, or something else -- maybe a modification of $\Sys$? Using a script of ``A'' was more intuitive when we were leavng $B$ and $C$ invariant.}
Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$. Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire, and still conserve the phenotype. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).

The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output. A minimal system's components are both reachable and observable. In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs. Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response. Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components. After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton. 
Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question. 
\js{should the sentence of evolution of nearly minimal be in the paper? Or should we save this for our next paper?}

A large, adapted population will move around in $\allS$ under environmental and selective stasis. In the minimal dimension neutral evolution proceeds as a stochastic coordinate transformation on the elements in $\Sys_{\min}$. Even if there is conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.

@@

  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).

@@

    For instance, the diploid $\{\Sys, \Sys' \}$ has system architecture $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$, the average of its two haploid systems. During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \! \times \! \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems (Appendix \ref{apx:recombination}). Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).

@@

\paragraph{Hybridization}
\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} 
The means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma^2$. 
\js{I think you mean $\sigma^{2}$? I changed it.}
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
\js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
These are depicted in figure \ref{fig:conceptual_fig}.

>>>>

\newcommand{\Sys}{\mathcal{S}}
\newcommand{\allS}{\mathcal{A}}
%\newcommand{\allS}{\mathcal{A}}

@@

\plr{Additional ideas to consider adding:}
\begin{itemize}
    %\item Haldane's rule (easy point in discussion: makes F1s look like F2s)
    %\item add point about F1: quartic and F2: quadratic to results and maybe abstract/discussion
    %\item hybrid vigor (need to do calculation)
    \item discuss linearity, linearization, and canalization in introduction
    %\item obtain estimates of variation in $A$ from thermodynamic occupancy model
\end{itemize}

%\plri{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}

%\plri{Need to come up with a consistent term for ``the $A_{ij}$''s. -- ``regulatory coefficients''? ``genotype''?}

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}
Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
This outlook necessitates the advancement of conceptual frameworks of such precision, 
only mathematics will suffice, as models allow the development of concrete numerical predictions. 
%Previously it has been suggested that any idealized study of evolution is incomplete 
%without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.
\plr{probably some more recent Wagner papers in this line as well?}
%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory. 
%
%Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
%
%
%It is commonly taught that an organism's genome contains the heritable material 
%that natural selection filters and that an organism's phenotype directly determines its
%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
%undrstood molecular machinery -- and it is a major goal common to many different disciplines
%within the life sciences to elucidate its form, function, and evolution. These aims are
%delicately intertwined and a comphrehensive understanding of a system's evolution requires
%an understanding of its function and vice versa.

@@

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
%\plr{not clear what you mean by "actually"}\js{I think we discussed it's clear enough for an intro?}
Models with different parameter schemes, or even different mechanics can be equally accurate, 
but still not \emph{actually} agree with what is being modelled. 
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
In chemical engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.  
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
In fact, evolutionary
biologists who study convergent versus parallel evolution, and developmental homology and analogy, are very familiar with such functional symmetries; 
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks. 
We focus primarily on the neutral scenario: 
that is where phenotype is conserved over evolutionary time. 
%however, this framework could be applied to a wider set of evolutionary scenarios. 

@@

-- that is we describe the set of all gene network architectures that yield identical phenotypes in equivalent environments, and show that all linear biological systems can, in principal, can undergo systems drift and rewire. 
Under neutrality, a population will drift through the set of all possible phenotypically equivalent linear gene networks.
Consequentially, two phenotypically equivlanet yet allopatrically isolated populations, will likely produce incompatibile hybrids, despite the absence of adaptation, directional selection, or environmental change.
Under these conditions, speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

@@

Such a biological system $\mathcal{S}$ is given by,
  \begin{align}\label{eqn:system}
    \mathcal{S} := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
  \end{align}

@@

%\plr{This paragraph feels like the introduction.}
%The literature is filled with detailed observations of molecular systems and their diversity. 
%\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
%There are examples of significant diversity in the networks underlying processes such as 
%circadian rhythm \citep{sancar2008intelligent}, 
%cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
%pattern formation \plr{cite?}, 
%and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
Despite a symmetry in functionality or phenotype systems can often differ, sometimes substantially, at the molecular level. 
How many different mechanisms have the same function? 
Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
%Any linear and minimal gene network 
%-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
%-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
%  \begin{equation}
%    \begin{aligned}
%      h(t) &= C e^{A t} B 
%      = C V^{-1} V e^{A t} V^{-1} V B \\
%      &= C V^{-1} e^{V A V^{-1} t} V B 
%      = \bar{C} e^{\bar{A} t} \bar{B}
%    \end{aligned}
%  \end{equation}
%Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
%have the same phenotype, for all possible inputs, if and only if they are related by a change of coordinates. 
%Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
%That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
%\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
%    i.e. behaviour across many possible inputs.}
%Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
%    VB &= B \implies \bar{B} = B \\
%    CV &= C \implies \bar{C} = C
%\plr{this sentence lacks an object (or some sentence part?)}
%In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.
%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
%Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VAV^{-1}, VB, CV )$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
 % where $V$ is any invertible matrix and it's elements are free parameters. 
%Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.
%The Kalman decomposition decomposes a system into \emph{reachable} and \emph{observable} components -- that is coefficients that respond to an input and coefficients that produce an observable output. A minimal system's components are both reachable and observable. In this form, it is possible to add components that respond to an input, but do not influence the output, or components that, in principle, influence the output, but do not respond to any inputs. Any system with such components is, by definition, non-minimal, and any system can be modified to include any finite number of extra unreachable and/or unobservable components, without modifying its impulse response. Decomposing the system allows for the easy manipulation of ``superfluous''(but not necessarily expendable) components. After unreachable/unobservable coefficients are added, changing coordinates ``stirs'' the system coefficients such that the previously superfluous coefficients can become essential to system funciton. 
%Whether or not evolution constrains biological systems to be minimal or nearly minimal is an open question.

\begin{definition}[Phenotypic equivalence of systems]
    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:system}
    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
    \textbf{phenotypically equivalent} 
    if
    \begin{align*}
        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
    \end{align*}
    Equivalently, this occurs if and only if
    \begin{align*}
        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
    \end{align*}
    where $h$ and $\bar h$ are the impulse responses of the two systems.
\end{definition}

One way to find other systems equivalent to a given one
is by change of coordinates:
if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
have the same dynamics because their impulse responses are equal:
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
%\begin{align*}
%    CV^{-1}( zI - VAV^{-1})^{-1}VB
%    =
%    CV^{-1}V( zI - A)^{-1}V^{-1}VB
%    =
%    C( zI - A)^{-1}B .
%\end{align*}
However, the converse is not necessarily true: 
systems can have identical impulse responses without being changes of coordinates of each other.
In fact, systems with identical impulse responses can involve interactions between different
numbers of molecular species, and thus be in different dimensions.

We refer to $\calA(\Sys_0)$ as the set of all systems equivalent to $\Sys_0$, regardless of dimension:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.

The set of all systems phenotypically equivalent to a given system $(A,B,C)$ 
is elegantly described using the Kalman decomposition,
which also clarifies the system dynamics.
To motivate this, first note that the input $u(t)$ only directly pushes the system
in directions lying in the span of the columns of $B$.
As a result, different combinations of input can 
move the system in any direction that lies in the \emph{reachable subspace},
which we denote by $\reachable$,
and is defined to be the closure of $\spn(B)$ under applying $A$
(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$).
Analogously to this, we define
the \emph{observable subspace}, $\mathcal{O}$,
to be the closure of $\spn(C^T)$ under applying $A$.
(Or: $\unobservable$ is the largest $A$-invariant subspace
contained in the null space of $C$;
and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)

\begin{theorem}[Kalman decomposition] \label{thm:kalman}
        For any system $(A,B,C)$ with corresponding Kalman basis matrix $P$,
        the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
        \begin{align*}
            \widehat A = PAP^{-1}
            &=
            \left[ \begin{array}{cccc}
                A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
                0 & A_{\ro} & 0 & A_{\ro,\nro} \\
                0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
                0 & 0 & 0 & A_{\nro}
            \end{array} \right] ,
        \end{align*}
        and
        \begin{align*}
            \widehat B = PB
            &=
            \left[ \begin{array}{cccc}
                B_{\rno} \\
                B_{\ro} \\
                0 \\
                0 
            \end{array} \right] ,
        \end{align*}
        and
        \begin{align*}
            \widehat C = CP^{-1}
            &=
            \left[ \begin{array}{cccc}
                0 & C_{\ro} & C_{\nrno} & 0 
            \end{array} \right] .
        \end{align*}
        The impulse response of both systems is given by
        \begin{align*}
          h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro} .
        \end{align*}
\end{theorem}

In the latter case, we say that the system is \emph{minimal} 
-- there is no equivalent system with a smaller number of molecular species.
Note that this says that any two equivalent minimal systems
are changes of coordinates of each other.

Since any system can be put into this form,
and once in this form, its impulse response is determined only by 
$C_{\ro}$, $A_{\ro}$, and $B_\ro$,
therefore, the set of all equivalent systems are parameterized by
the dimension $n$,
the choice of basis ($P$),
the remaining submatrices in $\widehat A$, $\widehat B$, and $\widehat C$
(which are unconstrained),
and a invertible transformation of $\spn(P_{\ro})$, which we call $T_\ro$.

%Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
%Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
%  \begin{align}
%    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
%  \end{align}
It is remarkable to note that even with the relationships between environment, kryptotype, and phenotype constant, and in the minimal dimension, there are still almost always degrees of freedom. These correspond to distinct genetic networks that perform indistinguishable functions. For example, the equivalent systems in Example \ref{ex:all_osc} are minimal, and share common B and C matrices.

It is also remarkable to note that there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, there is always more than one possible gene regulatory network architecture per phenotype. 
%This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).

For the remainder of the paper, we interpret $\allS$ as the phenotypially neutral landscape, wherein a large population will drift under environmental and selective stasis. Even if the phenotype is constrained and remains constant through evolutionary time, the molecular mechanism underpinning it is not constrained and likely will not be conserved.

@@

  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \times 2}$, and $(A,B,C) \in \Sys_{\min}$).

@@

    For instance, the diploid $\{\Sys, \Sys' \}$ has system architecture $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$, the average of its two haploid systems.
    %During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \! \times \! \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems (Appendix \ref{apx:recombination}). Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).

     During meiosis, a diploid recombines its two genomes to form a gamete by randomly swapping rows or columns between its two systems.
  \emph{E.g.} gamete systems $(A'', B'', C'')$, produced by the diploid $\{\Sys, \Sys'\}$, are formed by recombining two systems $\Sys = (A,B,C)$ and $\Sys' = (A', B', C')$  to form $\Sys''$ such that,
  %\begin{align*}
   $A'' = MA + (I-M)A'$, 
   $B'' = MB + (I-M)B'$, and
   $C'' = CM + C'(I-M)$.
 % \end{align*}
  $M$ is a diagonal matrix where each diagonal element is a Bernoulli random variable ($M_{ii} = 0$ or $1$ with equal probability, and $M_{ij}=0$ if $i \neq j$). If systems are different dimensions the smaller system elements can be augmented with $0$s (\emph{e.g.} $\left[ \begin{smallmatrix} A & 0 \\ 0 & 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} B \\ 0 \end{smallmatrix} \right], \left[ \begin{smallmatrix} C & 0 \end{smallmatrix}\right]$).

@@

\paragraph{Hybridization}
The means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma^2$. 
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
These are depicted in figure \ref{fig:conceptual_fig}.
